\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}

### 2016-09-23 chapters/background.tex
meta-bnf project description

\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.

@@

\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
Keeping the model of the code in the AST form has several advantages.
One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
The code is generated out of the always-valid AST beyond user's reach.

@@

\paragraph{PE4MPS.}
PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
It has tw components: PE parser and PE4MPS plugin for MPS.
Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
These extensions tell (guide, hint) the parser what the AST node layout should look like.
Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
The parser then uses this information when generating the projectional editor for am AST node.
Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.

@@

\paragraph{ANTLR{\_}MPS.}
ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
The author of this project is Fabien Campagne
The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.

@@

\paragraph{mps-metabnf.}
\todo{Popis na zaklade emailove diskuze (Nizozemci).}

>>>>

\JB{Navrhuji odstranit}
\PV{Souhlasim}
We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.

@@

\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
\PV{Spis bych zrusil tady, kapitola 1 mi prijde vhodnejsi pro zmineni motivace}
Keeping the model of the code in the AST form has several advantages.
One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
The code is generated out of the always-valid AST beyond user's reach.

@@

\paragraph{PE4MPS.}
PE4MPS~\cite{ref:PE4MPS} is a project by Marco Lombardo that is trying to solve the same problem as we do.
It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
It has tw components: PE parser and PE4MPS plugin for MPS.
Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
These extensions tell (guide, hint) the parser what the AST node layout should look like.
Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
The parser then uses this information when generating the projectional editor for am AST node.
Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.

@@

\paragraph{ANTLR{\_}MPS.}
\PV{Tenhle projekt byl z tech tri nejmene uzitecny a podle me i nejmene funkcni, tak mozna by se dal v ramci zkracovani vypustit. V dalsim projektu je na nej reference, tak treba ji vyhodit.}
ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
The author of this project is Fabien Campagne.
The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.

@@

\paragraph{mps-metabnf.}
\PV{Pridal jsem popis projektu}
% TODO - citace na https://github.com/DSLFoundry/mps-metabnf
The mps-metabnf project comes from the DSLFoundry group and also takes on grammar importing.
Even though it is in a very early stage of development, it holds some interesting ideas.
Similarly to ANTLR{\_}MPS, authors of mps-metabnf decided to create an MPS language describing the grammar.
When grammar is being imported into MPS, it is just transformed into the terms of this MPS language.
User is then able to adjust this grammar.
After this, a second step of the import process is needed, that generates the final MPS language out of this adjusted grammar definition.

The key takeaways from the mps-metabnf project are that the intermediate grammar MPS representation, when finished, will offer some powerful means of adjustment.
Using the projectional editor, the user can easily specify information about the code layout, export the target MPS language and in case further adjustments are needed, come back to the grammar definition and fix any problems.
The MPS language of the grammar definition can be easily extended with needed features (indenting, line breaks...), which is superior to our approach.

It also has some downsides.
Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
Another downside might be the need for a double transformation implementation -- first one being the import of the text grammar to the MPS grammar definition.
